---
title: "Lab 1 - BINF 6310"
author: "Erin Tsai"
date: "2/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 1
##### What is the mean and variance for the loaded dice?
```{r}
mean = (1*0.1) + (2*0.1) + (3*0.1) + (4*0.1) + (5*0.1) + (0.5*6)
variance = ((0.1)*(1-mean)^2) + ((0.1)*(2-mean)^2) + ((0.1)*(3-mean)^2) + ((0.1)*(4-mean)^2) + ((0.1)*(5-mean)^2) + ((0.5)*(6-mean)^2)
```

* The mean for the loaded die is **`r mean`**.
* The variance for the loaded die is **`r variance`**.

### Question 2
##### Make a function in R that "rolls" this die; return a vector containing the rolls.
```{r}
numRolls <- 3000
loaded_prob <- c(0.1,0.1,0.1,0.1,0.1,0.5)

rolls <- vector(length=numRolls, mode = "double")
for(i in 1: numRolls) rolls[i] <- sample(1:6, 1, replace = TRUE, prob = loaded_prob)
rolls
```

### Question 3
##### Make a histogram of some large number of rolls. Do the rolls of the loaded die approximate a uniform distribution?
* No, they do not approximate uniform distribution because 6 has a higher chance of being rolled than the other numbers. 
```{r, echo=FALSE}
hist(rolls)
```

### Question 4
##### Means vs. trial size plots of loaded die. How many rolls appear to be necessary to get convergence on the expected values for the mean and variance?
```{r}
mean_load <- mean(rolls)
var_load <- var(rolls)

trialSizes <- c(5,10,15,20,25,30,40,50,100,200,300,400,500,1000,2000,3000,4000,5000,10000,20000,30000,100000)
means <- vector(mode = "double", length=length(trialSizes))
variances <-vector(mode = "double", length=length(trialSizes))

for(i in 1:length(trialSizes))
{
    rolls1 <- vector(length = trialSizes[i], mode = "double")
    
    for(j in 1:trialSizes[i])
    {
      rolls1[j] <- sample(1:6, 1, replace = TRUE, prob = loaded_prob)
    }
    
    means[i] <- mean(rolls1);
    
    variances[i] <- var(rolls1)
}

plot(log10(trialSizes), means, main = "Means vs. Trial Size")
lines(log10(trialSizes), rep(mean_load, length(trialSizes)))
plot(log10(trialSizes), variances, main = "Variances vs. Trial Size")
lines(log10(trialSizes), rep(var_load, length(trialSizes)))
```




* It takes approximately **1000-3000 rolls** to converge on the expected mean.
* It takes approximately **1000-3000 rolls** to converge on the expected variance.


